Compact Core Galaxies in the RESOLVE Survey

### 2016-04-28 paperdraft.tex
adding new figure
lots of new text in intro

\noindent The term compact stellar system (CSS) encompasses a class of galaxies that spans the radius range ($\sim10 - 800$ pc), lying between globular clusters (GCs) and normal elliptical galaxies (Es). Different types of CSSs include ultra compact dwarf galaxies \citep[UCDs;][]{Phillipps2001} and compact elliptical galaxies \citep[cEs;][]{Faber1973}. There are many unanswered questions about how CSSs form and evolve through time. One fact that makes the study of these compact galaxies so intriguing is their apparent scarcity, which raises the question as to whether they are an intermediate and short-lived evolutionary phase in the life of a galaxy, or simply hard to find due to their size. Only in the last 15 years have we discovered enough ($\sim 40$) of these objects to enable studies of their formation scenarios and evolution over cosmic time.

In contrast, dwarf elliptical galaxies (dEs) and their nucleated dwarf elliptical galaxies (dE,Ns) are ...

UCDs, the smallest of the CSSs mentioned above with radii ranging from 10--100 pc \citep[e.g.,][]{Norris2014}, are largely believed to be either the high mass extension of the globular cluster population \citep{Drinkwater2000, Mieske2002} or tidally-stripped remnants of nucleated dwarf galaxies \citep{Bekki2001, Bekki2003, Jennings2015, Zhang2015}. \citet{Norris2011} find evidence that UCDs are a ``mixed bag'' of both giant GCs and stripped nuclei based on their analysis of UCDs discovered near the elliptical galaxy NGC3923 and the S0 galaxy NGC4546. \citet{Seth2014} find a supermassive black hole at the center of another UCD, providing further evidence of tidal stripping being common for UCDs.

Moving up in size, cEs have radii ranging typically from $\sim$100--600 pc \citep[e.g.,][]{Norris2014}. Similarly to UCDs, they are typically thought to be the result of a tidal stripping event, as is the case for the prototypical cE, M32 \citep{Choi2002, Graham2002}. The evidence for tidal stripping includes finding tidal streams near cEs \citep{SmithCastelli2008a,Chilingarian2009} and one cE, studied in \citet{Kormendy1997}, that is believed to host a central massive black hole. On the other hand, \citet{Kormendy2012a} argue that cEs could be the low mass extension to the normal E population created via dissipative formation in gas rich environments. \citet{Norris2014} argue that both scenarios are likely at play. %explain more

In contrast to CSSs, dEs and dE,Ns, are the largest of the CSSs mentioned with radii ranging from $\sim$600-800 pc \citep[e.g.,][]{Norris2014}. Similar again to UCDs and cEs, dEs are often thought to be either tidally stripped remnants (\citet{Crnojevi2014} find dEs in M31 that show signs of tidal stripping) or the low mass extension to elliptical galaxies \citep{Crawford2016}. They could also be the remnants of LCBDs that have been ram-pressure stripped as they entered the cluster environment and subsequently quenched.  There is some evidence that dE,Ns are the progenitors of UCDs (\citet{Pfeffer2013}, \citet{Zhang2015}, \citet{Liu2015}).

Clues to how CCGs form may be given by their morphology i.e., one- or two-component structure, environments, star formation histories, and kinematics. \citet{Pfeffer2013} predict that tidally stripped remnants will display multi-component surface brightness profiles due to lingering outer galaxy structures. Dissipative processes could also cause cores to form as gas rushes to the center and star formation begins. These effects could vary, however, based on the degree of stripping and how much time has passed since the stripping or dissipative formation occurred.

CSSs have been found in a range of environments thus far. In the cluster environment we find UCDs \citep{Price2009,Madrid2010,Hasegan2005,Jones2006,Mieske2009,Wehner2007,Misgeld2008}, cEs \citep{Chilingarian2007,SmithCastelli2012,Price2009}, and dEs \citep{SmithCastelli2012,Koo1994,Guzman1996,Crawford2016}. Fewer CSSs have been found in groups: \citet{Evstigneeva2007} find one definite UCD and two candidates in the Dorado group as well as two candidates in the NGC1400 group and \citet{Hau2009} find a UCD near the Sombrero galaxy; \citet{Huxor2011} and \citet{Chilingarian2010} find cEs in groups; and \citet{Crnojevi2014}  \citet{Penny2014} find dEs in groups. Lastly, only cEs have been found in isolated environments: \citet{Huxor2013}, \citet{Paudel2014}, and \citet{Chilingarian2015} find cEs separated from other galaxies by $>1$ Mpc. CSSs found in group or cluster environments may be tidally stripped remnants, whereas isolated CCSs may rather be disippatively formed like normal Es.

Another useful tool we have in discerning between formation scenarios is the galaxy's stellar velocity dispersion ($\sigma$). \citet{Bender1992} and \citet{Bekki2003} show that $\sigma$ will remain largely unchanged when tidal stripping occurs, even though the \Reff\ and stellar mass will be greatly reduced. Thus, galaxies that are offset to higher $\sigma$ in a Faber-Jackson (stellar mass -- $\sigma$) relation \citep{faber1976} are most likely tidally stripped, while galaxies falling along the relation are most likely the low-mass extension of normal Es.

Given the similarities in formation scenarios and environments for CSSs, we propose a study of these galaxies as a part of a continuum of compact galaxies instead of being broken up into (perhaps somewhat arbitrary) classes based on \Reff.  Critical to this work is a ubiquitous sample of CSSs that will allow us to analyze their formation as a function of all environments. Our goal is therefore to create a complete and volume-limited sample of compact galaxies with key data quantifying their demographics, environments, and kinematics. In this work we present the first complete, volume-limited sample of 96 compact galaxies that we term ``compact core galaxies'' or CCGs. We supplement this analysis with data on the colors and star formation histories of CCGs, which could point to recent star formation as a sign of stripping or disippative formation.

This paper is organized as follows. We describe our parent sample, the RESOLVE survey, in \autoref{resolve} and our comparison sample, the AIMSS catalog, in \autoref{aimss}. Our methods for performing seeing-deconvolved one and two component fits for the entire RESOLVE sample using \textsc{Galfit} are presented in \autoref{deconv}. We then detail the selection of the CCG sample from the RESOLVE survey and quantify our CCG demographics in \autoref{CCGs}. We next describe our photometric and spectroscopic data and analysis in \autoref{methods}. We present a statistical distribution of environments, star formation histories, and velocity dispersions for the complete sample in \autoref{results}. In \autoref{discussion}, we present a discussion of our results in the context of CCG formation and evolution. Lastly, the summary and future work are given in \autoref{conclusions}.

>>>>

\noindent The term compact stellar system (CSS) encompasses a class of galaxies that spans the radius range between globular clusters (GCs) and normal elliptical galaxies (Es). Different types of CSSs include compact elliptical galaxies \citep[cEs;][]{Faber1973} and ultra compact dwarf galaxies \citep[UCDs;][]{Phillipps2001}. There are many unanswered questions about how CSSs form and evolve through time. One fact that makes the study of these compact galaxies so intriguing is their apparent scarcity, which raises the question as to whether they are an intermediate and short-lived evolutionary phase in the life of a galaxy, or simply hard to find due to their size. Only in the last 15 years have we discovered enough ($\sim 40$) of these objects to enable studies of their formation scenarios and evolution over cosmic time.

CSSs have radii ranging typically from $\sim$10--600 pc \citep[e.g.,][]{Norris2014}, and there are multiple ideas for how they form. The smaller of the CSSs, UCDs may simply be the high mass extension of the GC population \citep{Drinkwater2000, Mieske2002}. The is also evidence that UCDs are created via the tidal stripping of nucleated dwarf galaxies \citep{Bekki2001, Bekki2003, Jennings2015, Zhang2015}. \citet{Seth2014} find a supermassive black hole at the center of another UCD, providing further evidence of tidal stripping being common for UCDs. Similarly, cEs are also thought to be the result of tidal stripping events, as argued for the prototypical cE, M32 \citep{Choi2002, Graham2002, Huxor2011}. More evidence is provided by the discovery of tidal streams near cEs \citep{SmithCastelli2008a,Chilingarian2009} and one cE, studied in \citet{Kormendy1997}, that is believed to host a central massive black hole. On the other hand, \citet{Wirth1984,Kormendy2009,Kormendy2012a} argue that cEs are the low mass extension to the normal E population created via dissipative formation in gas rich environments, such as mergers or in the two-phase formation process detailed in \citet{Oser2010}. One of the most successful recent searches for CSSs is the AIMSS catalog \citep{Norris2014}: with a wealth of newly discovered UCDs and cEs to examine, the authors argue that both the high-mass GCs and tidal stripping scenarios are likely for UCD formation and both tidal stripping and low-mass E scenarios are likely for cE formation.

In contrast to CSSs, dwarf elliptical galaxies (dEs) and their nucleated dwarf elliptical galaxies (dE,Ns) are quite common, especially in clusters. \citet{Binggeli1991} find that of 225 dwarf galaxies present in the Virgo cluster, 174 are classfied as dEs or dE,Ns based on their sufrace brightness profiles. dEs and dE,Ns are also larger than CSSs with radii ranging from $\sim$600-800~pc \citep[e.g.,][]{Norris2014}, which may perhaps help their numbers. Similar to cEs, dEs are often thought to be either tidally stripped remnants \citep{Crnojevi2014} or the low mass extension to elliptical galaxies \citep{Crawford2016}. They could also be the remnants of luminous compact blue dwarfs (LCBDs) that have been ram-pressure stripped and often quenched as they enter the cluster environment. There is also some evidence that dE,Ns are the progenitors of UCDs \citep{Pfeffer2013,Zhang2015,Liu2015}.

Given the similarities in formation scenarios between CSSs and dEs, we introduce the term ``compact core galaxy'' or ``CCG'' to encompass CSSs and dEs with radii up to 800~pc. The goal of this paper is to study these galaxies as a part of a continuum of compact galaxies in order to discern between formation scenarios. Clues to how CCGs form may be given by their morphology i.e., one- or two-component structure, environments, star formation histories, and kinematics. We highlight these differences below.

%structure
The morphologies of \citet{Pfeffer2013} predict that tidally stripped remnants will display multi-component surface brightness profiles due to lingering outer galaxy structures. *** M32 has a bulge plus disk profile (Graham 2002)*** Dissipative processes could also cause cores to form as gas rushes to the center and star formation begins. *** Huxor cE has no underlying disk *** These effects could vary, however, based on the degree of stripping and how much time has passed since the stripping or dissipative formation occurred.

%environments
CSSs have been found in a range of environments thus far. In the cluster environment we find UCDs \citep{Price2009,Madrid2010,Hasegan2005,Jones2006,Mieske2009,Wehner2007,Misgeld2008}, cEs \citep{Chilingarian2007,SmithCastelli2012,Price2009}, and dEs \citep{SmithCastelli2012,Koo1994,Guzman1996,Crawford2016}. Fewer CSSs have been found in groups: \citet{Evstigneeva2007} find one definite UCD and two candidates in the Dorado group as well as two candidates in the NGC1400 group and \citet{Hau2009} find a UCD near the Sombrero galaxy; \citet{Huxor2011} and \citet{Chilingarian2010} find cEs in groups; and \citet{Crnojevi2014}  \citet{Penny2014} find dEs in groups. Lastly, only cEs have been found in isolated environments: \citet{Huxor2013}, \citet{Paudel2014}, and \citet{Chilingarian2015} find cEs separated from other galaxies by $>1$ Mpc. CSSs found in group or cluster environments may be tidally stripped remnants, whereas isolated CCSs may rather be disippatively formed like normal Es.

*** include kicked out scenarios here ***

%colors

%kinematics
Another useful tool we have in discerning between formation scenarios is the galaxy's stellar velocity dispersion ($\sigma$). \citet{Bender1992} and \citet{Bekki2003} show that $\sigma$ will remain largely unchanged when tidal stripping occurs, even though the \Reff\ and stellar mass will be greatly reduced. Thus, galaxies that are offset to higher $\sigma$ in a Faber-Jackson (stellar mass -- $\sigma$) relation \citep{faber1976} are most likely tidally stripped, while galaxies falling along the relation are most likely the low-mass extension of normal Es.

In this paper, we present a volume-limited sample of CCGs with key data quantifying their morphologies, environments, star formation histories, and kinematics. This paper is organized as follows. We describe our parent sample, the RESOLVE survey, in \autoref{resolve} and our comparison sample, the AIMSS catalog, in \autoref{aimss}. Our methods for performing seeing-deconvolved one and two component fits for the entire RESOLVE sample using \textsc{Galfit} are presented in \autoref{deconv}. We then detail the selection of the CCG sample from the RESOLVE survey and quantify our CCG demographics in \autoref{CCGs}. We next describe our photometric, environmental, and spectroscopic data and analysis in \autoref{methods}. We present a statistical distribution of environments, star formation histories, and velocity dispersions for the complete sample in \autoref{results}. In \autoref{discussion}, we present a discussion of our results in the context of CCG formation and evolution. Lastly, the summary and future work are given in \autoref{conclusions}.
